A Tale of Two Dictionary Learning Problems

### 2020-08-05 manuscripts/thesis/results.tex
chapters

%\chapter{Results}
\section{Results}\label{Results}

@@

Precise statements of our results require that we first identify some combinatorial criteria on the supports\footnote{Recall that a vector $\mathbf{x}$ is said to be \emph{supported} in $S$ when $\mathbf{x} \in \text{\rmfamily span}\{\mathbf{e}_j: j\in S\}$, with $\mathbf{e}_j$ forming the standard column basis.} of sparse vectors. Let $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$.  A \emph{hypergraph} on vertices $[m]$  is simply any subset $\mathcal{H} \subseteq 2^{[m]}$. We say that $\mathcal{H}$ is \textit{$k$-uniform} when $\mathcal{H} \subseteq {[m] \choose k}$. The \emph{degree} $\deg_\mathcal{H}(i)$ of a node $i \in [m]$ is the number of sets in $\mathcal{H}$ that contain $i$, and we say $\mathcal{H}$ is \emph{regular} when for some $r$ we have $\deg_\mathcal{H}(i) = r$ for all $i$ (given such an $r$, we say $\mathcal{H}$ is \textit{$r$-regular}). We also write $2\mathcal{H} := \{ S \cup S': S, S' \in \mathcal{H}\}$.  The following class of structured hypergraphs is a key ingredient in this work.

@@

We next give a quantitative generalization of the spark condition (\ref{SparkCondition}) to combinatorial subsets of supports. The \emph{lower bound} of an $n \times m$ matrix $\mathbf{M}$ is the largest $\alpha$ with \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$ \cite{Grcar10}. By compactness of the unit sphere, every injective linear map has a positive lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then submatrices formed from $2k$ of its columns or less have strictly positive lower bounds.

The lower bound of a matrix is generalized below in (\ref{Ldef}) by restricting it to the spans of certain submatrices\footnote{See \cite{vidal2005generalized} for an overview of the related ``union of subspaces" model.} associated with a hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ of column indices. Let $\mathbf{M}_S$ denote the submatrix formed by the columns of a matrix $\mathbf{M}$ indexed by $S \subseteq [m]$ (setting $\mathbf{M}_\emptyset := \mathbf{0}$).  In the sections that follow, we shall also let $\bm{\mathcal{M}}_S$ denote the column-span of a submatrix $\mathbf{M}_S$, and $\bm{\mathcal{M}}_\mathcal{G}$ to denote $\{\bm{\mathcal{M}}_S\}_{S \in \mathcal{G}}$.  We define:  % Define $L_\mathcal{H}(\mathbf{M})$ as follows:
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \ falpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \mathbf{0} \neq \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
writing also $L_{k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{In compressed sensing literature, \mbox{$1 - \sqrt{k} L_k(\mathbf{M})$}  is the asymmetric lower restricted isometry constant for $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.\label{ripfootnote}}  As explained above, compactness implies that $L_{2k}(\mathbf{M}) > 0$ for all $\mathbf{M}$ satisfying \eqref{SparkCondition}. Clearly, $L_{\mathcal{H}'}(\mathbf{M}) \geq L_\mathcal{H}(\mathbf{M})$ whenever $\mathcal{H}' \subseteq \mathcal{H}$, and similarly any $k$-uniform $\mathcal{H}$ satisfying $\cup \mathcal{H} = [m]$ has $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ (letting $L_{2k}$ := $L_m$ whenever $2k > m$).

We are now in a position to state our main result, though for expository purposes we leave the quantity $C_1$ %(a function of $\mathbf{A}$, $\mathbf{x}_1, \ldots, \mathbf{x}_N$, and $\mathcal{H}$) 
undefined until Sec.~\ref{DUT}. All results  below assume real matrices and vectors.

@@

%\cite{li2004analysis, Georgiev05, Aharon06, Hillar15}
We should stress here that framing the problem in terms of hypergraphs has allowed us to show, unlike in previous research on the subject, that the matrix $\mathbf{A}$ need not necessarily satisfy \eqref{SparkCondition} to be recoverable from data. As an example, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ with $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$ and take $\mathcal{H}$ to be all consecutive pairs of indices $1, \ldots ,6$ arranged in cyclic order. Then for $k=2$, the matrix $\mathbf{A}$ fails to satisfy \eqref{SparkCondition} while still obeying the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} for dictionary recovery.%, hence guaranteeing \eqref{Cstable} . %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
%This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\overline x}_i$ is known to have a size that grows polynomially in $m$ and $k$.

A practical implication of Thm.~\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One need simply check that the matrix and codes satisfy the (computable) assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$. In general, however, there is no known efficient procedure. We defer a brief discussion on this point to the next section.
%Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.  
%. on $\mathbf{A}$ and the $\mathbf{x}_i$, respectively.

@@

@@

We claim that a similar phenomenon applies to datasets of vectors with a stable sparse representation. Briefly, following the same procedure as in \cite[Sec.~IV]{Hillar15}, for $k < m$ and $n \geq \min(2k, m)$, we may consider the ``symbolic'' dataset $Y = \{\mathbf{A}\mathbf{x}_1,\ldots,\mathbf{A} \mathbf{x}_N\}$ generated by an indeterminate $n \times m$ matrix $\mathbf{A}$ and $m$-dimensional $k$-sparse vectors $\mathbf{x}_1, \ldots, \mathbf{x}_N$ indeterminate within their supports, which form a regular hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ satisfying the SIP. Restricting \mbox{$(k-1){m \choose k} + 1$} indeterminate $\mathbf{x}_i$ to each support in $\mathcal{H}$, and letting $\textbf{M}$ be the $n \times N$ matrix with columns $\mathbf{A}\mathbf{x}_i$, it can be checked that when $f(\mathbf{M}) \neq 0$ for a substitution of real numbers for the indeterminates, all of the assumptions on $\mathbf{A}$ and the $\mathbf{x}_i$ in Thm.~\ref{DeterministicUniquenessTheorem} are satisfied. We therefore have the following.  % ; in particular, $\mathbf{A}$ satisfies \eqref{SparkCondition}

@@

We close this section with some comments on the optimality of our results.  The linear scaling for $\varepsilon$ in \eqref{epsdel} is essentially optimal (e.g., see \cite{arias2013fundamental}), but a basic open problem remains: how many samples are necessary to determine the sparse coding model? 
Our results demonstrate that sparse codes $\mathbf{x}_i$ drawn from only a polynomial number of $k$-dimensional subspaces permit stable identification of the generating dictionary $\mathbf{A}$. 
This lends some legitimacy to the use of the model in practice, where data in general are unlikely (if ever) to exhibit the exponentially many possible $k$-wise combinations of dictionary elements required by (to our knowledge) all previously published results.

Consequently, if $k$ is held fixed or if the size of the support set of reconstructing codes is polynomial in $\overline m$ and $k$, then a practical (polynomial) amount of data suffices to identify the dictionary.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, tillmann2015computational}. Even so, in the next section we discuss how probabilistic guarantees can in fact be made for any number of available samples.

>>>>

\chapter{Results}\label{Results}

\section{Introduction}

TODO

@@

\section{Combinatorial designs for sparse supports}

A precise statement of the results requires first the identification of some combinatorial criteria on the supports\footnote{Recall that a vector $\mathbf{x}$ is said to be \emph{supported} in $S$ when $\mathbf{x} \in \text{\rmfamily span}\{\mathbf{e}_j: j\in S\}$, with $\mathbf{e}_j$ forming the standard column basis.} of sparse vectors. Let $\{1, \ldots, m\}$ be denoted $[m]$, its power set $2^{[m]}$, and ${[m] \choose k}$ the set of subsets of $[m]$ of size $k$.  A \emph{hypergraph} on vertices $[m]$  is simply any subset $\mathcal{H} \subseteq 2^{[m]}$. Let us say that $\mathcal{H}$ is \textit{$k$-uniform} when $\mathcal{H} \subseteq {[m] \choose k}$. The \emph{degree} $\deg_\mathcal{H}(i)$ of a node $i \in [m]$ is the number of sets in $\mathcal{H}$ that contain $i$, and we say $\mathcal{H}$ is \emph{regular} when for some $r$ we have $\deg_\mathcal{H}(i) = r$ for all $i$ (given such an $r$, we say $\mathcal{H}$ is \textit{$r$-regular}). Let us also write $2\mathcal{H} := \{ S \cup S': S, S' \in \mathcal{H}\}$.  The following class of structured hypergraphs is a key ingredient in this work.

@@

Next, a quantitative generalization of the spark condition (\ref{SparkCondition}) to combinatorial subsets of supports is given. The \emph{lower bound} of an $n \times m$ matrix $\mathbf{M}$ is the largest $\alpha$ with \mbox{$\|\mathbf{M}\mathbf{x}\|_2 \geq \alpha\|\mathbf{x}\|_2$} for all $\mathbf{x} \in \mathbb{R}^m$ \cite{Grcar10}. By compactness of the unit sphere, every injective linear map has a positive lower bound; hence, if $\mathbf{M}$ satisfies \eqref{SparkCondition}, then submatrices formed from $2k$ of its columns or less have strictly positive lower bounds.

The lower bound of a matrix is generalized below in (\ref{Ldef}) by restricting it to the spans of certain submatrices\footnote{See \cite{vidal2005generalized} for an overview of the related ``union of subspaces" model.} associated with a hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ of column indices. Let $\mathbf{M}_S$ denote the submatrix formed by the columns of a matrix $\mathbf{M}$ indexed by $S \subseteq [m]$ (setting $\mathbf{M}_\emptyset := \mathbf{0}$).  In the sections that follow, let $\bm{\mathcal{M}}_S$ denote the column-span of a submatrix $\mathbf{M}_S$, and $\bm{\mathcal{M}}_\mathcal{G}$ to denote $\{\bm{\mathcal{M}}_S\}_{S \in \mathcal{G}}$. Define: % Define $L_\mathcal{H}(\mathbf{M})$ as follows:
%\begin{align*} 
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}(\mathbf{x}_1-\mathbf{x}_2)\|_2 }{ \sqrt{2k} \|\mathbf{x}_1-\mathbf{x}_2\|_2} : \mathbf{x}_1, \mathbf{x}_2 \in \cup_{S \in \mathcal{H}} \bm{\mathcal{M}}_S \right\},
%\end{align*} 
%
%where we write $L_{2k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$. Note that if $\mathcal{H}$ covers $[m]$, then $L_2 > L_\mathcal{H}$.\footnote{The reader should beware that $L_2 = L_{[m]}$, whereas $L = L_{\{[m]\}}$. For even $k$, the quantity $1 - \sqrt{k} L_k(\mathbf{M})$ is also known in the compressed sensing literature as the (asymmetric) lower restricted isometry constant \cite{Blanchard2011}.} Clearly, for any $\mathbf{M}$ satisfying \eqref{SparkCondition}, we have $L_{k'}(\mathbf{M}) > 0$ for  $k' \leq 2k$.
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{ \|\mathbf{M}_S\mathbf{x}\|_2 }{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H} \right\},
%\end{align*} 
%\begin{align*}
%L_\mathcal{H}(\mathbf{M}) := \max \left\{ \ falpha: \|\mathbf{M}_S\mathbf{x}\|_2 \geq \alpha \sqrt{k} \|\mathbf{x}\|_2 : S \in \mathcal{H}, \ \ \mathbf{x} \in \mathbb{R}^m \right\},
%\end{align*} 
\begin{align}\label{Ldef}
L_\mathcal{H}(\mathbf{M}) := \min \left\{ \frac{\|\mathbf{M}_S\mathbf{x}\|_2}{ \sqrt{k} \|\mathbf{x}\|_2} : S \in \mathcal{H}, \ \mathbf{0} \neq \mathbf{x} \in \mathbb{R}^{|S|} \right\},
\end{align} 
%
writing also $L_{k}$ in place of $L_\mathcal{H}$ when $\mathcal{H} = {[m] \choose k}$.\footnote{In compressed sensing literature, \mbox{$1 - \sqrt{k} L_k(\mathbf{M})$}  is the asymmetric lower restricted isometry constant for $\mathbf{M}$ with unit $\ell_2$-norm columns \cite{Blanchard2011}.\label{ripfootnote}}  As explained above, compactness implies that $L_{2k}(\mathbf{M}) > 0$ for all $\mathbf{M}$ satisfying \eqref{SparkCondition}. Clearly, $L_{\mathcal{H}'}(\mathbf{M}) \geq L_\mathcal{H}(\mathbf{M})$ whenever $\mathcal{H}' \subseteq \mathcal{H}$, and similarly any $k$-uniform $\mathcal{H}$ satisfying $\cup \mathcal{H} = [m]$ has $L_2 \geq L_{2\mathcal{H}} \geq L_{2k}$ (letting $L_{2k}$ := $L_m$ whenever $2k > m$).

\section{Deterministic uniqueness theorems}

We are now in a position to state the main result, though for expository purposes the quantity $C_1$ will be left%(a function of $\mathbf{A}$, $\mathbf{x}_1, \ldots, \mathbf{x}_N$, and $\mathcal{H}$) 
undefined until Sec.~\ref{DUT}. All results  below assume real matrices and vectors.

@@

%\cite{li2004analysis, Georgiev05, Aharon06, Hillar15}
It should be stressed here that framing the problem in terms of hypergraphs will allowed us to show, unlike in previous research on the subject, that the matrix $\mathbf{A}$ need not necessarily satisfy \eqref{SparkCondition} to be recoverable from data. As an example, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ with $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$ and take $\mathcal{H}$ to be all consecutive pairs of indices $1, \ldots ,6$ arranged in cyclic order. Then for $k=2$, the matrix $\mathbf{A}$ fails to satisfy \eqref{SparkCondition} while still obeying the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} for dictionary recovery.%, hence guaranteeing \eqref{Cstable} . %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
%This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\overline x}_i$ is known to have a size that grows polynomially in $m$ and $k$.

A practical implication of Thm.~\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One need simply check that the matrix and codes satisfy the (computable) assumptions of Thm.~\ref{DeterministicUniquenessTheorem} on $\mathbf{A}$ and the $\mathbf{x}_i$. In general, however, there is no known efficient procedure. A brief discussion on this point is deferred until later.
%Another practical implication of Thm.\ref{DeterministicUniquenessTheorem} is the following: there is an effective procedure sufficient to affirm if a proposed solution $(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{x}_N)$ to Prob.~\ref{InverseProblem} is indeed unique (up to noise and inherent ambiguities). One simply checks that $\mathbf{B}$ and the $\mathbf{\overline x}_i$ satisfy the respective assumptions of Thm.~\ref{DeterministicUniquenessTheorem}.  
%. on $\mathbf{A}$ and the $\mathbf{x}_i$, respectively.

@@

\section{Probabilistic uniqueness theorems}

@@

It turns out that a similar phenomenon applies to datasets of vectors with a stable sparse representation. Briefly, following the same procedure as in \cite[Sec.~IV]{Hillar15}, for $k < m$ and $n \geq \min(2k, m)$, we may consider the ``symbolic'' dataset $Y = \{\mathbf{A}\mathbf{x}_1,\ldots,\mathbf{A} \mathbf{x}_N\}$ generated by an indeterminate $n \times m$ matrix $\mathbf{A}$ and $m$-dimensional $k$-sparse vectors $\mathbf{x}_1, \ldots, \mathbf{x}_N$ indeterminate within their supports, which form a regular hypergraph $\mathcal{H} \subseteq {[m] \choose k}$ satisfying the SIP. Restricting \mbox{$(k-1){m \choose k} + 1$} indeterminate $\mathbf{x}_i$ to each support in $\mathcal{H}$, and letting $\textbf{M}$ be the $n \times N$ matrix with columns $\mathbf{A}\mathbf{x}_i$, it can be checked that when $f(\mathbf{M}) \neq 0$ for a substitution of real numbers for the indeterminates, all of the assumptions on $\mathbf{A}$ and the $\mathbf{x}_i$ in Thm.~\ref{DeterministicUniquenessTheorem} are satisfied. We therefore have the following.  % ; in particular, $\mathbf{A}$ satisfies \eqref{SparkCondition}

@@

\section{Discussion}

Let us conclude this chapter with some comments on the optimality of these results.  The linear scaling for $\varepsilon$ in \eqref{epsdel} is essentially optimal (e.g., see \cite{arias2013fundamental}), but a basic open problem remains: how many samples are necessary to determine the sparse coding model? 
These results demonstrate that sparse codes $\mathbf{x}_i$ drawn from only a polynomial number of $k$-dimensional subspaces permit stable identification of the generating dictionary $\mathbf{A}$. 
This lends some legitimacy to the use of the model in practice, where data in general are unlikely (if ever) to exhibit the exponentially many possible $k$-wise combinations of dictionary elements required by (to my knowledge) all previously published results.

Consequently, if $k$ is held fixed or if the size of the support set of reconstructing codes is polynomial in $\overline m$ and $k$, then a practical (polynomial) amount of data suffices to identify the dictionary.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, tillmann2015computational}. Even so, in the next chapter it will be discussed how probabilistic guarantees can in fact be made for any number of available samples.
